Epidemiological birth-death models with partner notification

### 2024-04-30 article/article.tex
Sign test

\section{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\citep{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classic epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

@@

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by~\citet{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen more than the others~\citep{Stadler2014}.

@@

@@

This model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without them or any of their induced cases being sampled:

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$. The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:

\begin{equation}
\scriptsize
\begin{split}
log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
 &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events, followed by child }\\
 & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
\end{split}
\end{equation}

The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g. by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}). To become identifiable it requires one of its parameters to be fixed. In practice, it is often the sampling probability, as it may be approximated from epidemiological data $\rho$ (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).

\subsection{MTBD models}
MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes. A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$. The time between events of the same type is hence modeled with exponential distribution.

@@

We now propose a non-Markovian extension of the BD model that adds partner notification (PN).  In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\rho_n$. Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\psi_p >> \psi$). BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_p >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate. This creates an observed tip in the transmission tree.
\end{itemize}

@@

The -PN extension can be easily generalized from BD to MTBD models: at the moment of sampling an individual of any type can notify their last partner with the probability $\rho_n$, who will then stop transmitting and get removed at a constant removal rate $\psi_n >> \psi_k \; \forall k \in \{1, \ldots, m\}$. The -PN extension therefore adds two parameters to an MTBD model: $\psi_p$ and $\rho_n$.

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). We give its algorithmic details in Appendix.

\section{PN test}~\label{sec:test}
To access whether a give tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two tips having a common parent).

The intuition behind the test is that in the presence of partner notification the tree will contain more cherries (two-tip clades) whose tips are close in time. Indeed, in a transmission tree internal nodes correspond to transmissions while tips correspond to sampling events. Hence if a transmission tree is complete, a cherry corresponds to a transmission from an infected individual to their most recent partner, followed by sampling of both. If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time. For an incomplete transmission tree (where some patients were not sampled), the only difference would be that some cherries would include hidden transmissions, and hence not correspond to the two most recent partners.

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

Let us assume there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$, our test calculates the difference between its tip times $d_i$. It then generates a reshuffled cherry $i'$: It picks 5 (adjustable parameter) cherries with the roots that are the closest in time to the root of $i$, randomly selects two tips among their tips, and calculates their time difference $d_{i'}$. The test then compares $d_{i'}$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal. Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries: $S' =  \sum\limits_{i,i'=1}^N sign(d_{i'} - d_i) \approx 0$. However, if some cherries are ``notified'' their real tip differences should be smaller that reshuffled ones, biasing the total sum: $S' > 0$.

The reason for which we are constructing a random cherry from the neighbouring instead of any cherries (in terms of root date), is that cherry branch lengths depend on the date of the cherry root. As tip sampling times are bounded from above by the end of the sampling period, the closer the cherry root is to the end of the sampling period, the shorter on average its branches will be (and less time difference they will be able to have in any case).

The reshuffled cherry tip difference generation is repeated 1\,000 times (adjustable parameter), hence obtaining 1\,000 array comparisons $S'^{(1)}, \ldots, S'^{(1\,000)}$.

Finally, the test reports the proportion of reshuffled arrays whose comparison was negative ($S'^{(k)} < 0$). Note that in the presence of partner notification, this proportion should be very low. 
The test therefore reports the probability of partner notification being present if the tree.

@@

\subsection{Performance on simulated data}~\label{sec:test:sim}
To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}, see Fig.~1 in \citet{Voznica2021} for a graphical representation of these models and their parameters), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models.

For each model, we generated 100 trees with 500--1\,000 tips with our MTBD/MTBD-PN tree simulator. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\psi_p}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\rho_n \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \rho_n$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\psi_p$ and $\rho_n$ values as the $i$-th trees under BD-PN and BDEI-PN models.

We then run the PN tests on each tree of these 6 datasets. For 100/99/100 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e. 100\%/99\%/100\% true negative and 0\%/1\%/0\% false positive results). For 97/99/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e. 97\%/99\%/96\% true positive and 3\%/1\%/4\% false negative results). These results are summarised in Table~\ref{tbl:pntest}. Our test therefore showed both high specificity and sensitivity.

\begin{table}[!h]\centering
\small
\caption{PN test performance on trees generated under BD, BDEI, BDSS, BD-PN, BDEI-PN and BDSS-PN models (100 trees per model). We report the number of trees for each the PN-test p-value was $<0.05$, $\geq 0.05$, and its mean p-value for each model dataset.}
\begin{tabular}{l|r|r|c}
 & \textbf{p-value $<0.05$} & \textbf{p-value $\geq0.05$} & \textbf{mean p-value} \\
  \midrule
\textbf{BD}& 0 & 100 & 0.424 \\
\textbf{BD-PN}& 97 & 3 & 0.004 \\
  \midrule
\textbf{BDEI}& 1 & 99 & 0.493 \\
\textbf{BDEI-PN}& 99 & 1 & 0.001 \\
  \midrule
\textbf{BDSS}& 0 & 100 & 0.455 \\
\textbf{BD-PN}& 96 & 4 & 0.004 \\
\end{tabular}
\label{tbl:pntest}
\end{table}

\section{Mathematical formulation of BD-PN model}
In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal). Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.  Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$. $p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch (with potentially different individuals at times $t$ and (the branch end) $t_i$).

In trees generated by the BD-PN model, however, two other types of branches are possible. Let us consider a tip corresponding to a person A who got sampled at time $t_A$ and successfully notified their partner B (see Fig.~\ref{fig:pn-branches}). According to our model assumption 2, notified individuals are always observed. Hence B must be sampled too (at time $t_B \geq t_A$) and correspond to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B. Therefore the individual on the path between AB and B cannot change (is always B) and only hidden transmissions (from B to someone else) where the recipient subtree stayed unobserved are possible along the branch(es) leading from AB to B (before $t_A$). Note, that non-hidden transmissions are also possible before the notification time $t_A$, and in such a case we know which subtree corresponds to the donor (the one that contains B). Between $t_A$ and $t_B$ no transmission is possible on the branch of B as, according to assumption 3, after notification notified individuals do not transmit further.

We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p^{(i,o)}(t_i) = 1\label{eq:p-o}
\end{cases}
\end{equation}

We will use an \textbf{partner probability once notified} $p^{(i,np)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i,np)}(t) = & \psi_p p^{(i,nt)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
p^{(i,o)}(t_i) = 1\label{eq:p-np}
\end{cases}
\end{equation}

Moreover, according to our model assumption 4, only the most recent partner can get notified. Hence, AB must be the parent node of A, and no hidden tranmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).

We will use a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for such cases (when no hidden transmission occurred along the branch):

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
\end{cases}
\end{equation}

Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B already got sampled via the standard procedure or was notified by someone else. In this case nothing changes for the AB-A branch, and the AB-B branch(es) stay oriented.

@@

First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified:

\begin{equation}
\scriptsize
l^{(u)}(i) = p^{i}(t_j) 2\lambda l^{(u)}(i_0)l^{(u)}(i_1) \label{eq:lu-int-int}
\end{equation}

Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:

>>>>

\section{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\citep{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\citep{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

@@

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by~\citet{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g. HIV) more than the others~\citep{Stadler2014}.

@@

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g. by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

@@

\subsection{MTBD models}
MTBD models~\citep{Stadler2013a} add population structure to the BD model by allowing different individual states, transmissions between them and state changes. A general MTBD model with $m$ individual states has $2m^2 + m$ parameters: An individual in state $k \in \{1, \ldots, m\}$ can be removed at a constant average rate $\psi_k$ (with pathogen sampling probability $\rho_k$), change their state to state $l \in \{1, \ldots, m\}$ at a constant average rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in state $l$ at a constant average rate $\lambda_{kl}$. The time between events % of the same type 
is hence modeled with exponential distribution.

@@

In this section we propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$. Upon notification, the partner is removed almost instantaneously (modeled via a constant notified removal rate $\phi >> \psi$). BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\phi >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate. This creates an observed tip in the transmission tree.
\end{itemize}

@@

The -PN extension can be generalized from BD to MTBD models: at the moment of sampling, an individual of any type can notify their last partner with the probability $\upsilon$, who will then stop transmitting and get removed at a constant removal rate $\phi >> \psi_k \; \forall k \in \{1, \ldots, m\}$. The -PN extension therefore adds two parameters to an MTBD model: $\phi$ and $\upsilon$.

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in Appendix.

\section{PN test}~\label{sec:test}
To access whether a given tree is generated in a presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).

The intuition behind the test is that in the presence of partner notification the tree will contain more cherries whose tips are close in time. Indeed, in a transmission tree internal nodes correspond to transmissions while tips correspond to sampling events. Hence if a transmission tree is complete, a cherry corresponds to a transmission from an infected individual to their most recent partner, followed by sampling of both. If there is no partner notification, these two sampling events are independent, while in the presence of partner notification the first sampling might trigger the second one, and they will be close in time. For an incomplete transmission tree (where some infected individuals were not sampled), the only difference would be that some cherries might include hidden transmissions.

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time. Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$. Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\citep{mooersBranchLengthsBirth2012}. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

Let us assume that there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$. We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $\{2k\}*$ gets swapped with the tip $\{2k-1\}*$ ($1 \leq k \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}. If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle (see Fig.~\ref{fig:tipswap}). We then calculate the time differences $d_{i'}$ between the tips of these reshuffles cherries. 
The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g. to month or even a year). Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries. However, if some cherries are ``notified'' their real tip differences should be systematically smaller than reshuffled ones. We therefore output the pvalue calculated by the sign test.

\begin{figure*}[tbhp]
\centering 
\includegraphics[width=0.8\textwidth]{Fig_cherryswap.png}
\caption{Reshuffled cherry generation. \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left). The randomly selected tip is shown with a bold branch for each cherry (e.g. the tip on the right for the leftmost cherry). The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$). \textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth. As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle. The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
\label{fig:tipswap} 
\end{figure*}

%The reason for which we are constructing a random cherry from the neighbouring instead of any cherries (in terms of root date), is that cherry branch lengths depend on the date of the cherry root. As tip sampling times are bounded from above by the end of the sampling period, the closer the cherry root is to the end of the sampling period, the shorter on average its branches will be (and less time difference they will be able to have in any case).

@@

\subsection{Performance on simulated data}~\label{sec:test:sim}
To illustrate the PN test performance, we applied it to the trees simulated under three classical MTBD-family models: BD, BDEI (Birth-Death Exposed-Infectious~~\citep{Stadler2014}) and BDSS (Birth-Death with SuperSpreading~\citep{Stadler2013a}), and under their -PN versions: BD-PN, BDEI-PN and BDSS-PN models. (See Fig.~1 in \citet{Voznica2021} for a graphical representation of BD, BDEI and DBSS models and their parameters).

For each model, we generated 100 trees with 500--1\,000 tips with our MTBD/MTBD-PN tree simulator. For each tree the parameter values were drawn randomly within the following bounds:
\begin{itemize}
\item $R_e = \frac{{\lambda}}{{\psi}} \in ]1, 5]$, 
\item removal rate $\psi \in ]1 / 20, 1 / 5]$,
\item sampling probability $\rho \in ]0.1, 0.9]$,
\item \textit{(for -PN models only)} ratio between the partner removal rate after notification and the removal rate $\frac{\phi}{\psi} \in ]50, 500]$,
\item \textit{(for -PN models only)} partner notification probability $\upsilon \in ]0.01/\rho, 0.9]$ (the lower bound was picked this way to ensure that the probability to be a notified partner is at least 0.01 $=\rho \upsilon$),
\item \textit{(for BDEI models only)} becoming-infectious rate (at which an exposed individual becomes infectious) $\mu \in ]0.2, 10]$,
\item \textit{(for BDSS models only)} fraction of superspreaders $f_{SS} = \frac{\lambda_{SS}}{\lambda_{SS} + \lambda_{SN}} \in ]0.05, 0.2]$,
\item \textit{(for BDSS models only)} superspreading transmission ratio $x_{SS} = \frac{\lambda_{SS}}{\lambda_{NS}} = \frac{\lambda_{SN}}{\lambda_{NN}} \in ]3, 10]$.
\end{itemize}

The common parameters were consistent between models, in a sense that the $i$-th tree under the BDSS model had the same $R_e$, $\psi$ and $\rho$ values as the $i$-th trees under BDSS-PN, BD(-PN) and BDEI(-PN) models; the $i$-th tree under the BDSS-PN model had the same $\phi$ and $\upsilon$ values as the $i$-th trees under BD-PN and BDEI-PN models.

We then run the PN tests on each tree of these 6 datasets. For 98/97/95 out of 100 trees in the BD/BDEI/BDSS dataset the PN test p-value was above $0.05$ (i.e. 98\%/97\%/95\% true negative and 2\%/3\%/5\% false positive results). For 95/95/96 out of 100 trees in the BD-PN/BDEI-PN/BDSS-PN dataset the PN test p-value was below $0.05$ (i.e. 95\%/95\%/96\% true positive and 5\%/5\%/4\% false negative results). These results are summarised in Table~\ref{tbl:pntest}. Our test therefore showed both high specificity and sensitivity.

\begin{table}[!h]\centering
\small
\caption{PN test performance on trees generated under BD, BDEI, BDSS, BD-PN, BDEI-PN and BDSS-PN models (100 trees per model). We report the number of trees for which the PN-test p-value was $<0.05$, $\geq 0.05$, and the mean p-value for each model dataset.}
\begin{tabular}{l|r|r|c}
 & \textbf{p-value $<0.05$} & \textbf{p-value $\geq0.05$} & \textbf{mean p-value} \\
  \midrule
\textbf{BD}& 2 & 98 & 0.479 \\
\textbf{BD-PN}& 95 & 5 & 0.015 \\
  \midrule
\textbf{BDEI}& 3 & 97 & 0.531 \\
\textbf{BDEI-PN}& 95 & 5 & 0.008 \\
  \midrule
\textbf{BDSS}& 5 & 95 & 0.505 \\
\textbf{BDSS-PN}& 96 & 4 & 0.008 \\
\end{tabular}
\label{tbl:pntest}
\end{table}

\section{Mathematical formulation of BD and BD-PN models}\label{math}

The BD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without this individual nor any of their induced cases being sampled:

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$. The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\citep{zhukovaFastAccurateMaximumLikelihood2022}:
In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree. % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

\begin{equation}
\scriptsize
\begin{split}
log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of tips} \\
 &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ transmission events, followed by child }\\
 & ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
\end{split}
\end{equation}

Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}). To become identifiable it requires one of its parameters to be fixed. In practice, it is often the sampling probability $\rho$, as it may be approximated from epidemiological data (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).

In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal). Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.  Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$. $p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch (with potentially different individuals at times $t$ and (the branch end) $t_i$).

In trees generated by the BD-PN model, however, two other types of branches are possible. Let us consider a tip corresponding to a person A who got sampled at time $t_A$ and successfully notified their partner B (see Fig.~\ref{fig:pn-branches}). According to our model assumption 2, notified individuals are always observed. Hence B must be sampled too (at time $t_B \geq t_A$) and corresponds to a tree tip, while the common ancestor of A and B (AB) corresponds to the transmission between A and B. Therefore the individual on the path between AB and B cannot change (is always B) and only hidden transmissions (from B to someone else) where the recipient subtree stayed unobserved are possible along the branch(es) leading from AB to B (before $t_A$). Note, that non-hidden transmissions are also possible before the notification time $t_A$, and in such a case we know which subtree corresponds to the donor (the one that contains B). Between $t_A$ and $t_B$ no transmission is possible on the branch of B as, according to assumption 3, after notification notified individuals do not transmit further.

We will use an \textbf{oriented (partner) branch probability} $p_o^{(i)}(t)$ for cases when the individual at the end of the branch (or considered time interval) is the same as at time $t$ but might have transmitted to someone else (unobserved) along the way:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_o^{(i)}(t) &=~~  \Big(\lambda + \psi\Big) p_o^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    &~~~ - \lambda p_o^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p_o^{(i)}(t_i) &=  1\label{eq:p-o}
\end{cases}
\end{equation}

We will use a \textbf{notified partner branch probability} $p_p^{(i)}(t)$ for cases when the individual at the end of the branch is the partner who got notified at time $t_n < t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_p^{(i)}(t) &=  \phi p_p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no sampling in the next infinitesimal time $\Delta t$ }\\
p_p^{(i)}(t_i) &=  1\label{eq:p-np}
\end{cases}
\end{equation}

Moreover, according to our model assumption 4, only the most recent partner can get notified. Hence, AB must be the parent node of A, and no hidden transmission is possible along the branch AB-A (as otherwise there would have been a partner of A who is more recent than B).

We will use a \textbf{notifier branch probability} $p_n^{(i)}(t)$ for such cases (when no hidden transmission occurred along the branch):

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_n^{(i)}(t) &=  \Big(\lambda + \psi\Big) p_n^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
p_n^{(i)}(t_i) &= 1\label{eq:p-nh}
\end{cases}
\end{equation}

Note that A could have also ``unsuccessfully'' notified B, if by the time $t_A$, when A decided to notify B, B had already got sampled via the standard procedure or got notified by someone else. In this case nothing changes for the AB-A branch, and the AB-B branch(es) stay oriented.

@@

\subsubsection*{Case 2.1.1: $i$ is an unnotified internal node with internal-node children $i0$ and $i1$}

First, let's assume that $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified (as $i$ is also unnotified):

\begin{equation}
\scriptsize
l^{(\not{n})}(i) = p^{(i)}(t_j) 2\lambda l^{(\not{n})}(i_0)l^{(\not{n})}(i_1) \label{eq:lu-int-int}
\end{equation}
